Lineage Analysis of CRISPR and shRNA Screens

### 2018-10-18 18Q3.Rmd

In a previous version of the shRNA screen, DEMETER ([GitHub](https://github.com/cancerdatasci/demeter)) was used to compute a dependency score for each gene by using the depletion values from each shRNA to infer the effect of target knockdown (on-target) and of expressing a given miRNA seed (off-target) in each cell line. More negative values indicate increased dependency, while more positive values indicate lower dependency. Zero represents the avergae dependency across all cell lines.

In the new data release, [DEMETER2](https://depmap.org/R2-D2/) ([GitHub repository](https://github.com/cancerdatasci/demeter2)), developed by McFarland et al. (2018) was used to analyze the Achilles screen. DEMETER2 expands on DEMETER by including parameters for cell-line-specific screen effects and noisy data, correcting for global differences in shRNA levels across cell lines, pooling data acorss cell lines via hierarchical modeling, and using Bayesian inference to compute uncertainty estimates. Now, a score of zero represents no dependency.

@@

@@

  plot <- ggplot(data = use_data, aes(x = Mutation_Status_Nonsilent, y = AUC)) +
    facet_wrap(~ Hugo_Symbol, drop = FALSE, ncol = length(genes_lvlA_filt)) +
    geom_hline(yintercept = 0, lty = 2, color = "darkgray") +
    geom_boxplot(mapping = aes(fill = Mutation_Status_Nonsilent), position = position_dodge(0.85)) +
    scale_fill_manual(values = use_color) +
    guides(color = FALSE) +
    geom_text(data = label_text, mapping = aes(x = 1.5, y = max(use_data$AUC), label = p.signif.adj)) +
    theme(legend.position = "top", axis.ticks.x = element_blank(), axis.text.x = element_blank(), axis.title.x = element_blank()) +
    labs(fill = "Mutation Status", y = "AUC", x = "Mutation Status", title = paste0(dataset, ": ", drug))
  return(plot)
}
```

@@

# GDSC
gdsc <- read.delim("./data_munging/data_drug_auc_gdsc_10_12_2018.csv", sep = "\t", header  = TRUE, row.names = 1, check.names = FALSE)
gdsc <- merge(drug_meta, gdsc, by = "CID")
gdsc_melt <- melt(gdsc, id.vars = c("CID", "Drug"), measure.vars = colnames(gdsc)[3:ncol(gdsc)], variable.name = "accession_id", value.name = "AUC")
gdsc_grid <- expand.grid("accession_id" = unique(gdsc_melt$accession_id), "Hugo_Symbol" = genes_lvlA)
gdsc_data <- merge(gdsc_grid, gdsc_melt, by = "accession_id", all.x = TRUE)
gdsc_data <- merge(gdsc_data, ccl_converter, by = "accession_id", all.x = TRUE)
gdsc_data <- merge(gdsc_data, maf_df_lvlA, by = c("Hugo_Symbol", "CCLE_Name", "Broad_ID"), all.x = TRUE)
gdsc_data$Mutation_Status_Nonsilent <- ifelse(is.na(gdsc_data$Mutation_Status_Nonsilent), "Other", gdsc_data$Mutation_Status_Nonsilent)
gdsc_data$Hugo_Symbol <- factor(gdsc_data$Hugo_Symbol)
gdsc_data$AUC <- as.numeric(gdsc_data$AUC)
gdsc_data$Color_Nonsilent <- ifelse(gdsc_data$Mutation_Status_Nonsilent == "Other", "paleturquoise3", "turquoise4")
gdsc_data$Color_Nonsilent <- factor(gdsc_data$Color_Nonsilent)
gdsc_data <- filter(gdsc_data, !is.na(gdsc_data$AUC))
saveRDS(gdsc_data, "./data_munging/rds/ctrp_data_18Q3.rds", compress = "xz")
```

@@

```{r fig.height=4, fig.width=20}
crispr_ptmuts_lvlA_filt <- filter(crispr_data_ptmuts, Hugo_Symbol %in% genes_lvlA_filt)
crispr_ptmuts_lvlA_filt$Hugo_Symbol <- ordered(crispr_ptmuts_lvlA_filt$Hugo_Symbol, levels = genes_lvlA_filt)
crispr_ptmuts_lvlA_filt_color <- as.character(crispr_ptmuts_lvlA_filt$Color_Nonsilent)
names(crispr_ptmuts_lvlA_filt_color) <- crispr_ptmuts_lvlA_filt$Mutation_Status_Nonsilent

@@

crispr_ptmuts_lvlA_filt_plot_faceted <- ggplot(data = crispr_ptmuts_lvlA_filt, aes(x = Mutation_Status_Nonsilent, y = Score)) +
  facet_wrap(~ Hugo_Symbol, drop = FALSE, ncol = length(genes_lvlA)) +
  geom_hline(yintercept = 0, lty = 2, color = "darkgray") +
  geom_boxplot(mapping = aes(fill = Mutation_Status_Nonsilent), position = position_dodge(0.85)) +
  scale_fill_manual(values = crispr_ptmuts_lvlA_filt_color) +
  guides(color = FALSE) +
  geom_text(data = label_text, mapping = aes(x = 1.5, y = 2, label = p.signif.adj)) +
  theme(legend.position = "top", axis.ticks.x = element_blank(), axis.text.x = element_blank(), axis.title.x = element_blank()) +
  labs(fill = "Mutation Status", y = "CERES Score", x = "Mutation Status")
crispr_ptmuts_lvlA_filt_plot_faceted
# ggsave("./plots_18Q3/manuscript/crispr_ptmuts_lvlA_filt_plot_faceted.png", crispr_ptmuts_lvlA_filt_plot_faceted, device = "png", dpi = 450, width = 20, height = 4, units = "in")
```

@@

```{r fig.height=4, fig.width=20}
ccle_ptmuts_lvlA_filt <- filter(ccle_data_ptmuts, Hugo_Symbol %in% genes_lvlA_filt)
ccle_ptmuts_lvlA_filt$Hugo_Symbol <- ordered(ccle_ptmuts_lvlA_filt$Hugo_Symbol, levels = genes_lvlA_filt)
ccle_drugs <- as.character(unique(ccle_ptmuts_lvlA_filt$Drug))
```

Plot code:
```{r, eval=FALSE}
ccle_lvlA_drugboxplots <- lapply(ccle_drugs, makeDrugBoxplots, dataset = "CCLE")
ccle_lvlA_drugboxplot_paths <- paste0(ccle_drugs, "_ccle_boxplots.png")
pwalk(list(ccle_lvlA_drugboxplot_paths, ccle_lvlA_drugboxplots), ggsave, path = "./plots_18Q3/manuscript/ccle_lvlA_boxplots", dpi = 450, width = 20, height = 4, units = "in")
```

@@

```{r fig.height=4, fig.width=20}
ctrp_ptmuts_lvlA_filt <- filter(ctrp_data_ptmuts, Hugo_Symbol %in% genes_lvlA_filt)
ctrp_ptmuts_lvlA_filt$Hugo_Symbol <- ordered(ctrp_ptmuts_lvlA_filt$Hugo_Symbol, levels = genes_lvlA_filt)
ctrp_drugs <- as.character(unique(ctrp_ptmuts_lvlA_filt$Drug))
```

Plot code:
```{r, eval=FALSE}
ctrp_lvlA_drugboxplots <- lapply(ctrp_drugs, makeDrugBoxplots, dataset = "CTRP")
ctrp_lvlA_drugboxplot_paths <- paste0(ctrp_drugs, "_ctrp_boxplots.png")
pwalk(list(ctrp_lvlA_drugboxplot_paths, ctrp_lvlA_drugboxplots), ggsave, path = "./plots_18Q3/manuscript/ctrp_lvlA_boxplots", dpi = 450, width = 20, height = 4, units = "in")
```

@@

```{r fig.height=4, fig.width=20}
gdsc_ptmuts_lvlA_filt <- filter(gdsc_data_ptmuts, Hugo_Symbol %in% genes_lvlA_filt)
gdsc_ptmuts_lvlA_filt$Hugo_Symbol <- ordered(gdsc_ptmuts_lvlA_filt$Hugo_Symbol, levels = genes_lvlA_filt)
gdsc_drugs <- as.character(unique(gdsc_ptmuts_lvlA_filt$Drug))
```

Plot code:
```{r, eval=FALSE}
gdsc_lvlA_drugboxplots <- lapply(gdsc_drugs, makeDrugBoxplots, dataset = "GDSC")
gdsc_lvlA_drugboxplot_paths <- paste0(gdsc_drugs, "_gdsc_boxplots.png")
pwalk(list(gdsc_lvlA_drugboxplot_paths, gdsc_lvlA_drugboxplots), ggsave, path = "./plots_18Q3/manuscript/gdsc_lvlA_boxplots", dpi = 450, width = 20, height = 4, units = "in")
```

@@

Match specific point mutations.
```{r fig.height=5, fig.width=20}
g2p_indications <- filter(read.delim("./data_munging/data_mutation_associations_appended.csv", sep = "\t", header = TRUE), Evidence.Level == "A")
maf_g2p_indications <- filter(maf_raw, Genome_Change %in% g2p_indications$MutationName)
crispr_g2p_indications <- merge(maf_g2p_indications, crispr_data_ptmuts, by = c("Hugo_Symbol", "CCLE_Name", "Broad_ID"))
dup_g2p_indications <- filter(crispr_g2p_indications[, c("Hugo_Symbol", "CCLE_Name", "Broad_ID")] %>% group_by(Hugo_Symbol, CCLE_Name, Broad_ID) %>% tally(), n > 1)
crispr_g2p_indications <- merge(crispr_g2p_indications, dup_g2p_indications, by = c("Hugo_Symbol", "Broad_ID"), all.x = TRUE)
crispr_g2p_indications <- filter(crispr_g2p_indications, is.na(n))

crispr_g2p_indications_plot <- ggplot(crispr_g2p_indications, aes(x = Hugo_Symbol, y = Score)) +
  geom_hline(yintercept = 0, lty = 2, color = "darkgray") +
  geom_boxplot(alpha = 0.5, color = "lightgray", outlier.shape = NA) +
  geom_jitter(width = 0.3, mapping = aes(color = group_general_lineage_name), size = 1) +
  theme_light() +
  theme(legend.title = element_blank()) +
  labs(x = "Gene", y = "CERES Score")
crispr_g2p_indications_plot

crispr_g2p_indications_muts_plot <- ggplot(crispr_g2p_indications, aes(x = Protein_Change, y = Score)) +
  facet_grid(~ Hugo_Symbol, scales = "free_x", space = "free") +
  geom_hline(yintercept = 0, lty = 2, color = "darkgray") +
  geom_boxplot(alpha = 0.5, color = "lightgray", outlier.shape = NA) +
  geom_jitter(width = 0.1, mapping = aes(color = group_general_lineage_name), size = 1) +
  theme_light() +
  theme(legend.title = element_blank(), axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "Gene", y = "CERES Score")
crispr_g2p_indications_muts_plot

>>>>

@@

```{r, include=FALSE, eval=FALSE}
# In a previous version of the shRNA screen, DEMETER ([GitHub](https://github.com/cancerdatasci/demeter)) was used to compute a dependency score for each gene by using the depletion values from each shRNA to infer the effect of target knockdown (on-target) and of expressing a given miRNA seed (off-target) in each cell line. More negative values indicate increased dependency, while more positive values indicate lower dependency. Zero represents the avergae dependency across all cell lines.
# 
# In the new data release, [DEMETER2](https://depmap.org/R2-D2/) ([GitHub repository](https://github.com/cancerdatasci/demeter2)), developed by McFarland et al. (2018) was used to analyze the Achilles screen. DEMETER2 expands on DEMETER by including parameters for cell-line-specific screen effects and noisy data, correcting for global differences in shRNA levels across cell lines, pooling data acorss cell lines via hierarchical modeling, and using Bayesian inference to compute uncertainty estimates. Now, a score of zero represents no dependency.
```

@@

  plot <- ggplot(data = use_data, aes(x = Mutation_Status_Nonsilent, y = AUC)) +
    facet_wrap(~ Hugo_Symbol, drop = FALSE, nrow = 2) +
               # ncol = length(genes_lvlA_filt)) +
    geom_boxplot(mapping = aes(fill = Mutation_Status_Nonsilent), position = position_dodge(0.85)) +
    scale_fill_manual(values = use_color) +
    guides(color = FALSE) +
    geom_text(data = label_text, mapping = aes(x = 1.5, y = max(use_data$AUC), label = p.signif.adj), nudge_y = 0.2) +
    theme(legend.position = "top", axis.ticks.x = element_blank(), axis.text.x = element_blank(), axis.title.x = element_blank()) +
    labs(fill = "Mutation Status", y = "AUC", x = "Mutation Status", title = paste0(dataset, ": ", drug))
  return(plot)
}
```

@@

# GDSC
gdsc <- read.delim("./data_munging/data_drug_auc_gdsc_10_12_2018.csv", sep = "\t", header  = TRUE, row.names = 1, check.names = FALSE)
gdsc <- merge(drug_meta, gdsc, by = "CID")
gdsc_melt <- melt(gdsc, id.vars = c("CID", "Drug"), measure.vars = colnames(gdsc)[3:ncol(gdsc)], variable.name = "accession_id", value.name = "AUC")
gdsc_grid <- expand.grid("accession_id" = unique(gdsc_melt$accession_id), "Hugo_Symbol" = genes_lvlA)
gdsc_data <- merge(gdsc_grid, gdsc_melt, by = "accession_id", all.x = TRUE)
gdsc_data <- merge(gdsc_data, ccl_converter, by = "accession_id", all.x = TRUE)
gdsc_data <- merge(gdsc_data, maf_df_lvlA, by = c("Hugo_Symbol", "CCLE_Name", "Broad_ID"), all.x = TRUE)
gdsc_data$Mutation_Status_Nonsilent <- ifelse(is.na(gdsc_data$Mutation_Status_Nonsilent), "Other", gdsc_data$Mutation_Status_Nonsilent)
gdsc_data$Hugo_Symbol <- factor(gdsc_data$Hugo_Symbol)
gdsc_data$AUC <- as.numeric(gdsc_data$AUC)
gdsc_data$Color_Nonsilent <- ifelse(gdsc_data$Mutation_Status_Nonsilent == "Other", "paleturquoise3", "turquoise4")
gdsc_data$Color_Nonsilent <- factor(gdsc_data$Color_Nonsilent)
gdsc_data <- filter(gdsc_data, !is.na(gdsc_data$AUC))
saveRDS(gdsc_data, "./data_munging/rds/gdsc_data_18Q3.rds", compress = "xz")
```

@@

```{r fig.height=7, fig.width=12}
crispr_ptmuts_lvlA_filt <- filter(crispr_data_ptmuts, Hugo_Symbol %in% genes_lvlA_filt)
crispr_ptmuts_lvlA_filt$Hugo_Symbol <- ordered(crispr_ptmuts_lvlA_filt$Hugo_Symbol, levels = genes_lvlA_filt)
crispr_ptmuts_lvlA_filt_color <- as.character(crispr_ptmuts_lvlA_filt$Color_Nonsilent)
names(crispr_ptmuts_lvlA_filt_color) <- crispr_ptmuts_lvlA_filt$Mutation_Status_Nonsilent

@@

crispr_ptmuts_lvlA_filt_plot_faceted <- ggplot(data = crispr_ptmuts_lvlA_filt, aes(x = Mutation_Status_Nonsilent, y = Score)) +
  facet_wrap(~ Hugo_Symbol, drop = FALSE, nrow = 2) +
             # ncol = length(genes_lvlA)) +
  geom_hline(yintercept = 0, lty = 2, color = "darkgray") +
  geom_boxplot(mapping = aes(fill = Mutation_Status_Nonsilent), position = position_dodge(0.85)) +
  scale_fill_manual(values = crispr_ptmuts_lvlA_filt_color) +
  guides(color = FALSE) +
  geom_text(data = label_text, mapping = aes(x = 1.5, y = 2, label = p.signif.adj), nudge_y = 0.1) +
  theme(legend.position = "top", axis.ticks.x = element_blank(), axis.text.x = element_blank(), axis.title.x = element_blank()) +
  labs(fill = "Mutation Status", y = "CERES Score", x = "Mutation Status")
crispr_ptmuts_lvlA_filt_plot_faceted
# ggsave("./plots_18Q3/manuscript/crispr_ptmuts_lvlA_filt_plot_faceted.png", crispr_ptmuts_lvlA_filt_plot_faceted, device = "png", dpi = 450, width = 20, height = 4, units = "in")
```

@@

```{r fig.height=4, fig.width=20}
ccle_ptmuts_lvlA_filt <- filter(ccle_data_lvlA, Hugo_Symbol %in% genes_lvlA_filt)
ccle_ptmuts_lvlA_filt$Hugo_Symbol <- ordered(ccle_ptmuts_lvlA_filt$Hugo_Symbol, levels = genes_lvlA_filt)
ccle_drugs <- as.character(unique(ccle_ptmuts_lvlA_filt$Drug))
```

Plot code:
```{r, eval=FALSE}
ccle_lvlA_drugboxplots <- lapply(ccle_drugs, makeDrugBoxplots, dataset = "CCLE")
ccle_lvlA_drugboxplot_paths <- paste0(ccle_drugs, "_ccle_boxplots.png")
pwalk(list(ccle_lvlA_drugboxplot_paths, ccle_lvlA_drugboxplots), ggsave, path = "./plots_18Q3/manuscript/ccle_lvlA_boxplots", dpi = 450, width = 12, height = 7, units = "in")
```

@@

```{r fig.height=4, fig.width=20}
ctrp_ptmuts_lvlA_filt <- filter(ctrp_data_lvlA, Hugo_Symbol %in% genes_lvlA_filt)
ctrp_ptmuts_lvlA_filt$Hugo_Symbol <- ordered(ctrp_ptmuts_lvlA_filt$Hugo_Symbol, levels = genes_lvlA_filt)
ctrp_drugs <- as.character(unique(ctrp_ptmuts_lvlA_filt$Drug))
```

Plot code:
```{r, eval=FALSE}
ctrp_lvlA_drugboxplots <- lapply(ctrp_drugs, makeDrugBoxplots, dataset = "CTRP")
ctrp_lvlA_drugboxplot_paths <- paste0(ctrp_drugs, "_ctrp_boxplots.png")
pwalk(list(ctrp_lvlA_drugboxplot_paths, ctrp_lvlA_drugboxplots), ggsave, path = "./plots_18Q3/manuscript/ctrp_lvlA_boxplots", dpi = 450, width = 12, height = 7, units = "in")
```

@@

```{r fig.height=4, fig.width=20}
gdsc_ptmuts_lvlA_filt <- filter(gdsc_data_lvlA, Hugo_Symbol %in% genes_lvlA_filt)
gdsc_ptmuts_lvlA_filt$Hugo_Symbol <- ordered(gdsc_ptmuts_lvlA_filt$Hugo_Symbol, levels = genes_lvlA_filt)
gdsc_drugs <- as.character(unique(gdsc_ptmuts_lvlA_filt$Drug))
```

Plot code:
```{r, eval=FALSE}
gdsc_lvlA_drugboxplots <- lapply(gdsc_drugs, makeDrugBoxplots, dataset = "GDSC")
gdsc_lvlA_drugboxplot_paths <- paste0(gdsc_drugs, "_gdsc_boxplots.png")
pwalk(list(gdsc_lvlA_drugboxplot_paths, gdsc_lvlA_drugboxplots), ggsave, path = "./plots_18Q3/manuscript/gdsc_lvlA_boxplots", dpi = 450, width = 12, height = 7, units = "in")
```

@@

Match specific point mutations.
```{r fig.height=5, fig.width=20, eval=FALSE}
g2p_indications <- filter(read.delim("./data_munging/data_mutation_associations_appended.csv", sep = "\t", header = TRUE), Evidence.Level == "A")
maf_g2p_indications <- filter(maf_raw, Genome_Change %in% g2p_indications$MutationName)
crispr_g2p_indications <- merge(crispr_data_ptmuts, maf_g2p_indications, by = c("Hugo_Symbol", "CCLE_Name", "Broad_ID"), all.x = TRUE)
crispr_g2p_indications$Mutation_Status_Nonsilent <- ifelse(is.na(crispr_g2p_indications$Mutation_Status_Nonsilent), "Other", crispr_g2p_indications$Mutation_Status_Nonsilent)
dup_g2p_indications <- filter(crispr_g2p_indications[, c("Hugo_Symbol", "CCLE_Name", "Broad_ID")] %>% group_by(Hugo_Symbol, CCLE_Name, Broad_ID) %>% tally(), n > 1)
crispr_g2p_indications <- merge(crispr_g2p_indications, dup_g2p_indications, by = c("Hugo_Symbol", "CCLE_Name", "Broad_ID"), all.x = TRUE)
crispr_g2p_indications$Genome_Change <- factor(crispr_g2p_indications$Genome_Change, levels = unique(crispr_g2p_indications$Genome_Change))
crispr_g2p_indications_nonNA <- filter(crispr_g2p_indications, is.na(n))

crispr_g2p_indications_plot <- ggplot(crispr_g2p_indications, aes(x = Hugo_Symbol, y = Score)) +
  geom_hline(yintercept = 0, lty = 2, color = "darkgray") +
  geom_boxplot(alpha = 0.5, color = "lightgray", outlier.shape = NA) +
  geom_jitter(width = 0.3, mapping = aes(color = Mutation_Status_Nonsilent), size = 1) +
  theme_light() +
  theme(legend.title = element_blank()) +
  labs(x = "Gene", y = "CERES Score")
crispr_g2p_indications_plot

crispr_g2p_indications_muts_plot <- ggplot(crispr_g2p_indications, aes(x = Protein_Change, y = Score)) +
  facet_grid(~ Hugo_Symbol, scales = "free_x", space = "free") +
  geom_hline(yintercept = 0, lty = 2, color = "darkgray") +
  geom_boxplot(alpha = 0.5, color = "lightgray", outlier.shape = NA) +
  geom_jitter(width = 0.1, mapping = aes(color = Mutation_Status_Nonsilent), size = 1) +
  theme_light() +
  theme(legend.title = element_blank(), axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "Gene", y = "CERES Score")
crispr_g2p_indications_muts_plot
